---
title: "Nimble - Poisson Regression"
output:
  html_document:
    theme: flatly
    df_print: paged
  pdf_document: default
---

```{r setup,message=FALSE}
library(knitr)
library(kableExtra)
library(MASS)                 # Ship data
library(nimble, warn.conflicts = FALSE)
library(coda)                 # MCMC diagnostics
library(mcmcplots)            # MCMC plot
```

## Poisson Regression with offset

-   $Y_i|X_i \sim Poisson(\mu_i)$

-   $log(\mu_i)=\eta_i=\beta_0+\beta_1x_{1i}+log(t_i)=(\beta_0+log(t_i))+\beta_1x_{1i}$

-   $\mu_i=exp(\eta_i)=exp(\beta_0+\beta_1x_{1i}+log(t_i))=exp(\beta_0+\beta_1x_{1i})t_i$

```{r}
##############
# Simulation #
##############
rm(list = ls())
set.seed(1)
n <- 1000                                   # number of observations
x <- rnorm(n)                               # explanatory variables
X <- cbind(1,x)                             # design matrix
true_betas <- c(0.5,1.5)                    # coefficients

t <- sample(1:10,n,replace=T)               # offset

eta <- X %*% true_betas+log(t)              # plus offset term
mu  <- exp(eta)
y   <- rpois(n,mu) 

# histogram
tmp<-table(y)/n*100
barplot(tmp,col="skyblue",ylab="Percent",xlab="Count")

```

Setting

| Parameter | True |
|:---------:|:----:|
| $\beta_0$ | 0.5  |
| $\beta_1$ | 1.5  |

### Frequentist Approach

```{r}
# Frequentist approach
m <- glm(y~x,family = "poisson",offset=log(t)) # allow intercept
summary(m)

```

## NIMBLE Procedure

### Step 1: Build Model

-   Write the Poisson model:

-   $Y_i|X_i \sim Poisson(\mu_i)$

-   $log(\mu_i)=\eta_i=\beta_0+\beta_1x_{1i}+log(t_i)$; or

-   $\mu_i=exp(\eta_i)=exp(\beta_0+\beta_1x_{1i}+log(t_i))=exp(\beta_0+\beta_1x_{1i})t_i$

```{r}
#######################
# Model specification #
#######################

# use bugs code in nimbleCode

code <- nimbleCode({
  for(i in 1:n) {
    y[i] ~ dpois(mu[i])
    #mu[i] <- exp(beta0+beta1*x[i]+log(t[i]))
    log(mu[i])<- beta0+beta1*x[i]+log(t[i])
  }
  beta0 ~ dnorm(0, sd = 100)
  beta1 ~ dnorm(0, sd = 100)
})

constants <- list(n = n)
data <- list(y = y,x = x, t=t)              # Need to add offset data
inits <- list(beta0=0, beta1 = 0)
# nimbleModel: check code specification is correct
model <- nimbleModel(code, constants = constants, data = data, inits = inits)
```

### Step 2: Configure Model

-   If we want to calculate WAIC, specify **enableWAIC = TRUE** in configuration.

-   Default sampler: random walk sampler (scalar adaptive Metropolis-Hastings sampling with a normal proposal distribution)

```{r}
modelConf <- configureMCMC(model, enableWAIC = TRUE,print = TRUE)
modelConf$printSamplers()      # Defalut in RW sampler
```

### Step 3: Compile Model

-   After specifying model, initial values can be changed. Don't need to run function **nimbleModel()** again.

```{r}
Cmodel <- compileNimble(model)
```

### Step 4: Build MCMC

```{r}
modelMCMC <- buildMCMC(modelConf)
modelMCMC <- compileNimble(modelMCMC, project = model)
```

### Step 5: Run MCMC

```{r}
niter <- 1500
burn  <- 500
set.seed(1)
samples <- runMCMC(modelMCMC, niter = niter,nburnin = burn, nchains = 1,
                   summary = TRUE, WAIC = TRUE)

mcmc.out <- samples$samples
head(mcmc.out)
```

```{r}
out<-samples$summary
kable(out,digits = 4, format = "html") %>% kable_styling(full_width = F)
```

```{r}
mcmc.out <- samples$samples
target <- c("beta0","beta1")
mbeta <- colMeans(mcmc.out)[target]
qbeta <- apply(mcmc.out,2,quantile,c(0.025,0.975))[,target]

```

```{r echo=FALSE}
tmp2 <- data.frame(true=true_betas,
                   est = mbeta,
                   lci=qbeta[1,],
                   uci=qbeta[2,])
row.names(tmp2) <- c("$\\beta_{0}$","$\\beta_{1}$")
kable(tmp2,digits = 4, format = "html",
      col.names = c("True",
                    "MCMC Estimate",
                    "Lower CI",
                    "Upper CI")) %>%
                     kable_styling(full_width = F)
```

-   Show WAIC

```{r}
# Show WAIC
samples$WAIC
```

-   Plot trace plot

```{r}
par(mfrow = c(1, 2), mai = c(.6, .4, .1, .2))
plot(mcmc.out[ , "beta0"], type = "l", xlab = "iteration",
     ylab = expression(beta[1]),col="green")
abline(h=mean(mcmc.out[ , "beta0"]))
plot(mcmc.out[ , "beta1"], type = "l", xlab = "iteration",
     ylab = expression(beta[2]),col="green")
abline(h=mean(mcmc.out[ , "beta1"]))
```

-   Plot MCMC information (traceplot, density plot,...rtc) from MCMC package.

```{r,eval=FALSE}
mcmcplot(mcmc.out,filename = "output-Poi")
```

# Case Study

-   Ship data from MASS package

-   Outcome ($Y$): Number of damage incidence.

-   Variables ($X$): Type (A-E), year of construction (60-64, 65-69, 70-74, 75-79).

-   Offset variable: aggregate months of service. The longer ship service, the higher incidence happens.

```{r}
data(ships)
ships$yearbuilt <- as.factor(ships$year)
fit_ships_poisson <- glm(incidents ~ type + yearbuilt,
                         offset = log(service), family = poisson,
                         data = ships, subset = (service != 0))
summary(fit_ships_poisson)
```

```{r}
d <- subset(ships,service != 0)
n  <- nrow(d)              # num of observations 
y  <- d$incidents          # outcome
x1 <- 1*(d$type=="B")      # dummy coding: ship type (ref="A")
x2 <- 1*(d$type=="C")
x3 <- 1*(d$type=="D")
x4 <- 1*(d$type=="E")
x5 <- 1*(d$year==65)     # period (ref="60")
x6 <- 1*(d$year==70)
x7 <- 1*(d$year==75)
t  <- d$service
X  <- cbind(x1,x2,x3,x4,x5,x6,x7)
p  <- ncol(X)              # number of parameters (not include intercept)
```

### Nimble

```{r}
#######################
# Model specification #
#######################

# use bugs code in nimbleCode

code <- nimbleCode({
  for(i in 1:n) {
    y[i] ~ dpois(mu[i])
    log(mu[i]) <- beta0+inprod(X[i,1:p],beta[1:p])+log(t[i])}

  beta0 ~ dnorm(0,sd=100)
  for (j in 1:p){beta[j] ~ dnorm(0,sd=100)}
})

constants <- list(n = n,p=p)
data <- list(y = y,X = X, t=t)
inits <- list(beta0=0, beta=rep(0,p))
# nimbleModel: check code specification is correct
model <- nimbleModel(code, constants = constants, data = data, inits = inits)
```

```{r}
modelConf <- configureMCMC(model, print = TRUE, enableWAIC = TRUE)

modelConf$printSamplers()

Cmodel <- compileNimble(model)

modelMCMC <- buildMCMC(modelConf)
modelMCMC <- compileNimble(modelMCMC, project = model)


niter <- 2000
burn  <- 1000
set.seed(1)
samples <- runMCMC(modelMCMC, niter = niter,nburnin = burn, nchains = 1,
                   WAIC = TRUE, summary = TRUE)
mcmc.out <- samples$samples
```

```{r}
out<-samples$summary
rownames(out)<-c("typeB","typeC","typeD","typeE","yearbuilt65","yearbuilt70","yearbuilt75","Intercept")
kable(out,digits = 4, format = "html") %>% kable_styling(full_width = F)
```
