---
title: "Nimble - Simple Linear Regression"
output:
  html_document:
    theme: flatly
    df_print: paged
  pdf_document: default
---

```{r setup,message=FALSE}
library(knitr)
library(kableExtra)
library(nimble, warn.conflicts = FALSE)
library(coda)                 # MCMC diagnostics
library(mcmcplots)            # MCMC plot
```

## Simple Linear Regression (SLR)

-   $y_i=\beta_0+\beta_1x_i+e_i$, where $e_i \stackrel{i.i.d}{\sim} N(0,\sigma^2)$
-   number of observation = 150

### Simulation Setting

```{r}
##############
# Simulation #
##############
set.seed(1)
p <- 1                                      # number of explanatory variables
n <- 150                                    # number of observations
x <- rnorm(n)                               # explanatory variables
X <- cbind(1,x)                             # design matrix
true_betas <- c(0.5,1.5)                    # coefficients
sigma  <- 1                                 # standard deviation
sigma2 <- sigma^2                           # variance
y <- rnorm(n, X %*% true_betas, sigma)      # outcome
```

Setting

| Parameter  | True |
|:----------:|:----:|
| $\beta_0$  | 0.5  |
| $\beta_1$  | 1.5  |
| $\sigma^2$ |  1   |

### Frequentist Approach

```{r}
# Frequentist approach
m <- lm(y~x) # allow intercept
summary(m)
```

## NIMBLE Procedure

### Step 1: Build Model

#### Step 1.1: write model in winbugs format

-   Use **nimbleCode()** function to write **BUGS** code.

-   Remember **{ }** inside the parentheses

```{r}
#######################
# Model specification #
#######################

# use bugs code in nimbleCode
code <- nimbleCode({
  for(i in 1:n) {y[i] ~ dnorm(beta0+beta1*x[i], sd = sigma)}
  beta0 ~ dnorm(0, sd = 100)
  beta1 ~ dnorm(0, sd = 100)
  sigma ~ dunif(0, 100)  # prior for variance components based on Gelman (2006)
  tau <- 1/pow(sigma,2)  # power function in BUGS
})
```

#### Step 1.2: give constants, data, and inits

```{r}
constants <- list(n = n)                   # constant index (# of obs/variables)
data <- list(y = y,x = x)                  # data
inits <- list(beta0=0, beta1=0,sigma=0.5)  # initial value
```

#### Step 1.3: verify your model

-   Use nimbleModel() function to verify your model.

-   If model is correct, you will have the message below.

-   Sometime, it can run for a while if your model is complicated. (e.g: hierarchical model)

```{r}
# nimbleModel: check code specifcation is correct
model <- nimbleModel(code, constants = constants, data = data, inits = inits)
```

### Step 2: Configure Model

-   Tell you which variables we are monitoring.

-   Which samplers (RW, Gibbs sampling, etc. ) we use for each parameter?

-   If nimble cannot detect conjugate posterior, the default sampler is RW (Metropolis-Hastings sampling with a normal proposal distribution.).

```{r}
mcmcConf <- configureMCMC(model)

mcmcConf$printSamplers() ## Look at sampler assignments in this case.
```

### Step 3: Compile Model

-   Use compileNimble () function for model object.

-   Compile in C++ for faster execution

```{r}
Cmodel <- compileNimble(model)
```

### Step 4: Build MCMC

-   Build MCMC with your configuration object with buildMCMC() function

-   Compile MCMC object again!!!

```{r}
modelMCMC <- buildMCMC(mcmcConf)
modelMCMC <- compileNimble(modelMCMC, project = model)
```

### Step 5: Run MCMC

-   Two ways to run the MCMC procedure.

-   First way: runMCMC() function

-   Second way: nimbleMCMCcode() function

#### Step 5.1(first way): runMCMC() function

-   Personally use the first one runMCMC(), but I will show nimbleMCMC() very quickly.

-   Remember to set the seed before run the MCMC.

```{r}
niter <- 1500
burn  <- 0
set.seed(1)
samples <- runMCMC(modelMCMC, niter = niter,nburnin = burn, nchains = 1,
                   summary = TRUE)
head(samples$samples)

```

-   Extract samples

    -   Calculate posterior means and $95\%$ credible intervals.

    -   Show trace plot

    -   Show ACF plot

```{r}
out<-samples$summary
kable(out,digits = 4, format = "html") %>% kable_styling(full_width = F)
```

```{r}
mcmc.out<-samples$samples
mbeta <- colMeans(mcmc.out)[c("beta0","beta1")]
qbeta <- apply(mcmc.out,2,quantile,c(0.025,0.975))[,c("beta0","beta1")]

msigma <- colMeans(mcmc.out)["sigma"]
qsigma <- apply(mcmc.out,2,quantile,c(0.025,0.975))[,"sigma"]
```

```{r echo=FALSE}
tmp2 <- data.frame(true=c(true_betas,sigma),
                   est = c(mbeta,msigma),
                   lci=c(qbeta[1,],qsigma[1]),
                   uci=c(qbeta[2,],qsigma[2]))
row.names(tmp2) <- c("$\\beta_{0}$","$\\beta_{1}$","$\\sigma$")
kable(tmp2,digits = 4, format = "html",
      col.names = c("True",
                    "MCMC Estimate",
                    "Lower CI",
                    "Upper CI")) %>%
                     kable_styling(full_width = F)
```

```{r}
par(mfrow = c(1, 3), mai = c(.6, .4, .1, .2))
plot(mcmc.out[ , "beta0"], type = "l", xlab = "iteration",
     ylab = expression(beta[1]),col="green")
abline(h=mean(mcmc.out[ , "beta0"]))
plot(mcmc.out[ , "beta1"], type = "l", xlab = "iteration",
     ylab = expression(beta[2]),col="green")
abline(h=mean(mcmc.out[ , "beta1"]))
plot(mcmc.out[ , "sigma"], type = "l", xlab = "iteration",
     ylab = expression(sigma),col="green")
abline(h=mean(mcmc.out[ , "sigma"])) # note: should put burin 
```

```{r}
par(mfrow = c(1, 3), mai = c(.6, .4, .1, .2))
acf(mcmc.out[, "beta0"])  
acf(mcmc.out[, "beta1"]) 
acf(mcmc.out[, "sigma"]) 
```

-   Plot MCMC information (trace plot, density plot,...etc) from MCMC package.

```{r,eval=FALSE}

# Specify a dirction to save HTML
dirfil <- "C:\\Users\\chech\\OneDrive - Medical University of South Carolina\\Programs\\Github\\Bayesian-Nimble\\example\\mc-output"
mcmcplot(mcmc.out,dir=dirfil,filename = "output-SLR")
```

-   **NOTE: sigma based on RW sampler need more burn-in period.**

```{r, message=FALSE}
niter <- 1500
burn  <- 500
set.seed(1)
samples <- runMCMC(modelMCMC, niter = niter,nburnin = burn, nchains = 1)
```

```{r, echo=FALSE}
msigma <- colMeans(samples)["sigma"]
qsigma <- apply(samples,2,quantile,c(0.025,0.975))[,"sigma"]

par(mfrow = c(1, 2), mai = c(.6, .4, .1, .2))
plot(samples[ , "sigma"], type = "l", xlab = "iteration",
     ylab = expression(sigma),col="green")
abline(h=mean(samples[ , "sigma"])) # note: should put burin 
acf(samples[, "sigma"]) 
```

#### Step 5.2(second way): nimbleMCMC() function

```{r}
mcmc.out2 <- nimbleMCMC(code = code, constants = constants,
                       data = data, inits = inits,
                       nchains = 1, niter = 1500,
                       summary = TRUE, WAIC = TRUE,
                       monitors = c('beta0','beta1','sigma'))


names(mcmc.out2) # "samples" "summary" ""WAIC

# MCMC estimate 
mcmc.out2$summary
# Model fitting
mcmc.out2$WAIC  
```

#### 

## Conjugate Posterior for $\sigma^2$ or $\tau$

-   $y_i|x_i,\beta_0,\beta_1,\tau \sim N( \beta_0+\beta_1x_i,\tau^{-1})$ and $\tau\sim Gamma(a,b)$

-   $\tau~\sim Gamma(a+n/2,b+\sum_{i=1}^{n}(y_i-\beta_0-\beta_1x_i)^2/2)$

```{r}
# Gamma prior for variance: GA(0.001,0.001) 
code <- nimbleCode({
  for(i in 1:n) {y[i] ~ dnorm(beta0+beta1*x[i], tau = tau)} 
  # if want to have conjugacy, need to specify tau (not sd)
  beta0 ~ dnorm(0, sd = 100)
  beta1 ~ dnorm(0, sd = 100)
  tau   ~ dgamma(shape = 0.001, rate = 0.001)
  var   <- 1/tau
  sigma <- sqrt(var)
})

```

-   Use Inverse-Gamma to update variance

```{r, eval=FALSE}
###########
# NOT RUN #
###########
# Inverse gamma for precision: Inv-ga(1,1)
code <- nimbleCode({
  for(i in 1:n) {y[i] ~ dnorm(beta0+beta1*x[i], var = var)} 
  # if want to have conjugacy, need to specify var
 beta0 ~ dnorm(0, sd = 100)
  beta1 ~ dnorm(0, sd = 100)
  var   ~ dinvgamma(shape = 0.0001, rate = 0.0001) # non-informative prior of 
                                                   # inverse gamma (1,1)
  tau   <- 1/var
  sigma <- sqrt(var)
})

```

-   Check the conjugacy for precision and variance

```{r}
constants <- list(n = n, p = p, x = x)
data <- list(y = y)
inits <- list(beta0=0, beta1 = 0, tau = 1)
# nimbleModel: check code specifcation is correct
model <- nimbleModel(code, constants = constants, data = data, inits = inits)
# Model configuration
modelConf <- configureMCMC(model, print = TRUE)
modelConf$printSamplers()               # show conjugacy

```

```{r}

modelConf$addMonitors(c("var","sigma")) # add monitor of var & sigma

# Compile model
Cmodel <- compileNimble(model)
Cmodel$beta0
# Biuld MCMC
modelMCMC <- buildMCMC(modelConf)
modelMCMC <- compileNimble(modelMCMC, project = model)

########
# MCMC #
########
niter <- 1500
burn  <- 0
set.seed(1)
samples <- runMCMC(modelMCMC, niter = niter,nburnin = burn, nchains = 1,
                   summary = T)
```

```{r}
out<-samples$summary
kable(out,digits = 4, format = "html") %>% kable_styling(full_width = F)
```

```{r}
mcmc.out<-samples$samples
par(mfrow = c(2, 3), mai = c(.6, .4, .1, .2))
plot(mcmc.out[ , "beta0"], type = "l", xlab = "iteration",
     ylab = expression(beta[1]),col="green")
abline(h=mean(mcmc.out[ , "beta0"]))
plot(mcmc.out[ , "beta1"], type = "l", xlab = "iteration",
     ylab = expression(beta[2]),col="green")
abline(h=mean(mcmc.out[ , "beta1"]))
plot(mcmc.out[ , "tau"], type = "l", xlab = "iteration",
     ylab = expression(sigma),col="green")
abline(h=mean(mcmc.out[ , "tau"])) 

acf(mcmc.out[, "beta0"])  
acf(mcmc.out[, "beta1"]) 
acf(mcmc.out[, "tau"]) 
```

### Conjugate ('Gibbs') Samplers

-   The current release of NIMBLE supports conjugate sampling of the relationships

|        Prior        | Sampling Distribution | Parameter |
|:-------------------:|:---------------------:|:---------:|
|        Beta         |       Bernoulli       |   prob    |
|                     |       Binomial        |   prob    |
|                     |   Negative Binomial   |   prob    |
|      Dirichlet      |      Multinomial      |   prob    |
|                     |      Categorical      |   prob    |
|        Flat         |        Normal         |   mean    |
|                     |       Lognormal       |  meanlog  |
|        Gamma        |        Poisson        |  lambda   |
|                     |        Normal         |    tau    |
|                     |       Lognormal       |  taulog   |
|                     |         Gamma         |   rate    |
|                     |     Inverse Gamma     |   scale   |
|                     |      Exponential      |   rate    |
|                     |  Double Exponential   |   rate    |
|                     |        Weibull        |  lambda   |
|       Normal        |        Normal         |   mean    |
|                     |       Lognormal       |  meanlog  |
| Multivariate Normal |  Multivariate Normal  |   mean    |
|       Wishart       |  Multivariate Normal  |   prec    |
|   Inverse Wishart   |  Multivariate Normal  |    cov    |

## Coda Package with Nimble

-   Effective sample size

-   Geweke Diagnostic

-   Gelman-Rubin Diagnostic

```{r}
# Effective Sample Sizes
  effectiveSize(mcmc.out[500:1000,"beta1"])
  effectiveSize(mcmc.out[500:1000,"tau"])

```

```{r}
# Geweke Diagnostics
  geweke.diag(mcmc.out[500:1000,"beta1"])$z
  2*pnorm(geweke.diag(mcmc.out[500:1000,"beta1"])$z)      
```

```{r}
# Gelman-Rubin (two chains)
  niter <- 1500
  burn  <- 0
  set.seed(1)
  samples <- runMCMC(modelMCMC, niter = niter,nburnin = burn, nchains = 2)
  mcmc.out.c1 <- samples$chain1
  mcmc.out.c2 <- samples$chain2
  
  names(samples)
  
  tauc1 <- mcmc(mcmc.out.c1[501:1500,"tau"])
  tauc2 <- mcmc(mcmc.out.c2[501:1500,"tau"])
  
  taulist <- mcmc.list(tauc1,tauc2)
  gelman.diag(taulist,autoburnin = F)
```
