---
title: "Nimble - Logistic Regression"
output:
  html_document:
    theme: flatly
    df_print: paged
  pdf_document: default
---

```{r setup,message=FALSE}
library(knitr)
library(kableExtra)
library(nimble, warn.conflicts = FALSE)
library(coda)                 # MCMC diagnostics
library(mcmcplots)            # MCMC plot
```

## Logistic Regression

-   $Y_i|x_i \sim Bernoulli(\pi_i)$

-   $logit(\pi_i)=\eta_i=\beta_0+\beta_1x_{1i}+\beta_2x_{2i}$

-   $\pi_i=P(y_i=1)=\frac{exp(\eta_i)}{1+exp(\eta_1)}$

```{r}
##############
# Simulation #
##############
rm(list = ls())
set.seed(1)
n  <- 500                              # number of observations
x1 <- rnorm(n)                         # explanatory variables
x2 <- rbinom(n,1,0.5)
X  <- cbind(1,x1,x2)                   # design matrix
k  <- ncol(X)                          # number of variables
true_betas <- c(1,1.5,-2)              # coefficients

eta  <- X %*% true_betas               # linear predictors
prob <- exp(eta)/(1+exp(eta))         
y    <- rbinom(n,1,prob)               # data
#table(y)/n                            # check proportion of 1's 
```

Setting

| Parameter | True |
|:---------:|:----:|
| $\beta_0$ | 1.0  |
| $\beta_1$ | 1.5  |
| $\beta_2$ |  -2  |

### Frequentist Approach

```{r}
# Frequentist approach
m <- glm(y~x1+x2,family = "binomial") # allow intercept
summary(m)
```

## NIMBLE Procedure

### Step 1: Build Model

-   Two ways to write logistic model and probit model.

-   $Y_i|x_i \sim Bernoulli(\pi_i)$

-   $logit(\pi_i)=\beta_0+\beta_1x_{1i}+\beta_2x_{2i}$; or

-   $\pi_i=ilogit(\beta_0+\beta_1x_{1i}+\beta_2x_{2i})$; equivalently $\pi_i=\frac{exp(\beta_0+\beta_1x_{1i}+\beta_2x_{2i})}{1+exp(\beta_0+\beta_1x_{1i}+\beta_2x_{2i})}$

-   $\Phi^{-1}(\pi_i)=\beta_0+\beta_1x_{1i}+\beta_2x_{2i}$; or

-   $\pi_i=\Phi(\beta_0+\beta_1x_{1i}+\beta_2x_{2i})$

-   NOTE: Data is generated from binomial distribution with logit link so probit model won't show the true $\beta$s.

```{r}
#######################
# Model specification #
#######################
code <- nimbleCode({
  for(i in 1:n) {
     y[i] ~ dbin(prob=p[i],size=1)
    ### 1. logit model ###
    logit(p[i]) <- inprod(beta[1:k],X[i,1:k])   # logit fn
    #p[i] <- ilogit(inprod(beta[1:k],X[i,1:k])) # inverse logit fn
  
    ### 2. probit model ###
    #probit(p[i]) <- inprod(beta[1:k],X[i,1:k]) # probit fn
    #p[i] <- iprobit(inprod(beta[1:k],X[i,1:k]))         # inverse probit fn
  }
    
  # specify prior for all betas
  for (j in 1:k){beta[j]~dnorm(0,sd=100)}  
})

constants <- list(n = n, k=k)     # constants
data <- list(y = y, X=X)          # data
inits <- list(beta=rep(0,k))      # initial values
# nimbleModel: check code specification is correct
model <- nimbleModel(code, constants = constants, data = data, inits = inits)
```

### Step 2: Configure Model

-   If we want to calculate WAIC, specify **enableWAIC = TRUE** in configuration.

-   Default sampler: random walk sampler (scalar adaptive Metropolis-Hastings sampling with a normal proposal distribution)

```{r}
modelConf <- configureMCMC(model, enableWAIC = TRUE,print = TRUE)
modelConf$printSamplers()      # Defalut in RW sampler
```

### Step 3: Compile Model

-   After specifying model, initial values can be changed. Don't need to run function **nimbleModel()** again.

```{r}
Cmodel <- compileNimble(model)
Cmodel$beta[1]<-1
Cmodel$beta[2]<--1
Cmodel$beta[3]<-0
```

### Step 4: Build MCMC

```{r}
modelMCMC <- buildMCMC(modelConf)
modelMCMC <- compileNimble(modelMCMC, project = model)
```

### Step 5: Run MCMC

```{r}
niter <- 1500
burn  <- 0
set.seed(1)
samples <- runMCMC(modelMCMC, niter = niter,nburnin = burn, nchains = 1,
                   summary = TRUE, WAIC = TRUE)

mcmc.out <- samples$samples
head(mcmc.out)

# length(apply(mcmc.out,2,unique))
```

```{r}
out<-samples$summary
kable(out,digits = 4, format = "html") %>% kable_styling(full_width = F)
```

```{r}
mcmc.out <- samples$samples
target <- c("beta[1]","beta[2]","beta[3]")
mbeta <- colMeans(mcmc.out)[target]
qbeta <- apply(mcmc.out,2,quantile,c(0.025,0.975))[,target]

```

```{r echo=FALSE}
tmp2 <- data.frame(true=true_betas,
                   est = mbeta,
                   lci=qbeta[1,],
                   uci=qbeta[2,])
row.names(tmp2) <- c("$\\beta_{1}$","$\\beta_{2}$","$\\beta_{3}$")
kable(tmp2,digits = 4, format = "html",
      col.names = c("True",
                    "MCMC Estimate",
                    "Lower CI",
                    "Upper CI")) %>%
                     kable_styling(full_width = F)
```

-   Show WAIC

```{r}
# Show WAIC
samples$WAIC
```

-   Plot MCMC information (traceplot, density plot,...rtc) from MCMC package.

```{r,eval=FALSE}
mcmcplot(mcmc.out,filename = "output-log")
```

## Change Samplers for $\beta$s

-   Default sampler: random walk sampler (scalar adaptive Metropolis-Hastings sampling with a normal proposal distribution)

-   check **help(samplers),** In help windows, show every sampler we can use to update parameters **(do not include "sampler\_")**. [[Link](https://www.r-bloggers.com/2017/05/better-block-sampling-in-mcmc-with-the-automated-factor-slice-sampler/)]

-   Suggestions: before adding samplers, removing the samplers first!

```{r,eval=FALSE}
help(samplers)
```

### 1.type="RW_block"

-   Adaptive Metropolis-Hastings algorithm with a multivariate normal proposal distribution (Roberts and Sahu, 1997)

```{r}
RW_modelConf <- configureMCMC(model, enableWAIC = TRUE,print = TRUE)
RW_modelConf$removeSamplers()
RW_modelConf$addSampler(c("beta[1]","beta[2]","beta[3]"),type="RW_block")
RW_modelConf$printSamplers()
```

```{r,echo=FALSE,message=FALSE,warning=FALSE}
Cmodel <- compileNimble(model)

modelMCMC <- buildMCMC(RW_modelConf)
modelMCMC <- compileNimble(modelMCMC, project = model)

set.seed(1)
samples <- runMCMC(modelMCMC, niter = niter,nburnin = burn, nchains = 1,
                   summary = TRUE, WAIC = TRUE)
out<-samples$summary
kable(out,digits = 4, format = "html",caption="RW_ block Sampler") %>% kable_styling(full_width = F)
```

### 2. type="slice"

-   Cannot use slice sampler on more than one target node.

```{r}
SLICE_modelConf <- configureMCMC(model, enableWAIC = TRUE,print = TRUE)
SLICE_modelConf$removeSamplers(c("beta[3]"))
SLICE_modelConf$addSampler(c("beta[3]"),type="slice")
SLICE_modelConf$printSamplers()
```

```{r,echo=FALSE,message=FALSE,warning=FALSE}
Cmodel <- compileNimble(model)

modelMCMC <- buildMCMC(SLICE_modelConf)
modelMCMC <- compileNimble(modelMCMC, project = model)

set.seed(1)
samples <- runMCMC(modelMCMC, niter = niter,nburnin = burn, nchains = 1,
                   summary = TRUE, WAIC = TRUE)
out<-samples$summary
kable(out,digits = 4, format = "html",,caption="Slice Sampler") %>% kable_styling(full_width = F)
```

### 3. type="AF_slice"

-   Automated factor slice sampler conducts a slice sampling algorithm on one or more model nodes.

```{r}
AF_SLICE_modelConf <- configureMCMC(model, enableWAIC = TRUE,print = TRUE)
AF_SLICE_modelConf$removeSamplers()
AF_SLICE_modelConf$addSampler(c("beta[1]","beta[2]","beta[3]"),type="AF_slice")
AF_SLICE_modelConf$printSamplers()
```

```{r,echo=FALSE,message=FALSE,warning=FALSE}
Cmodel <- compileNimble(model)

modelMCMC <- buildMCMC(AF_SLICE_modelConf)
modelMCMC <- compileNimble(modelMCMC, project = model)

set.seed(1)
samples <- runMCMC(modelMCMC, niter = niter,nburnin = burn, nchains = 1,
                   summary = TRUE, WAIC = TRUE)
out<-samples$summary
kable(out,digits = 4, format = "html",
      caption="Automated Factor Slice Sampler") %>% kable_styling(full_width = F)
```
