---
title: "Random Intercept Model"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: 72
---

# Model Specification 

-   See Lecture 23 Page 8

-   Observation level: Let $Y_{ij}$ be the outcome of subject $i$
    ($i=1,\ldots,n$) at $j$-th ($j=1,\ldots,n_i$) time:
    $$Y_{ij}=\beta_0+\beta_1t_{ij}+b_{i}+e_{ij},$$ where
    $t_{ij}\in \{1,2,\ldots,T\}$. We assume
    $b_{i}\overset{\text{iid}}{\sim}\text{N}(0,\sigma^2_b)$,
    $e_{ij}\overset{\text{iid}}{\sim}\text{N}(0,\sigma^2_e)$

-   Combining all observations:
    $$\boldsymbol{Y}=\boldsymbol{X}\boldsymbol{\beta}+\boldsymbol{Z}\boldsymbol{b}+\boldsymbol{e},$$
    where
    $\boldsymbol{Y}=(Y_{11},Y_{12},\ldots,Y_{1n_1},\ldots,Y_{n1},Y_{n2},\ldots,Y_{nn_n})_{N\times1}^T$
    and $\boldsymbol{b}=(b_1,\ldots,b_n)_{n\times1}^T$

-   If we write out the equation above: $$\begin{bmatrix}
    Y_{11}\\
    Y_{12}\\
    \vdots\\
    Y_{1n_1}\\
    \vdots\\
    Y_{n1}\\
    Y_{n2}\\
    \vdots\\
    Y_{nn_n}
    \end{bmatrix}_{N\times1}=\begin{bmatrix}
    1 & t_{11}\\
    1 & t_{12}\\
    \vdots &\vdots\\
    1 &t_{1n_1}\\
    \vdots &\vdots\\
    1 &t_{n1}\\
    1 &t_{n2}\\
    \vdots &\vdots\\
    1 & t_{nn_n}
    \end{bmatrix}_{N\times2}\begin{bmatrix} \beta_0  \\ \beta_1 \end{bmatrix}_{2\times1}+\begin{bmatrix}
    b_1\\
    b_1\\
    \vdots\\
    b_1\\
    \vdots\\
    b_n\\
    b_n\\
    \vdots\\
    b_n
    \end{bmatrix}_{N\times1}+\begin{bmatrix}
    e_{11}\\
    e_{12}\\
    \vdots\\
    e_{1n_1}\\
    \vdots\\
    e_{n1}\\
    e_{n2}\\
    \vdots\\
    e_{nn_n}
    \end{bmatrix}_{N\times1}$$

```{r setup,warning=FALSE,message=FALSE}
library(mcmcplots)      
library(nimble)
library(lme4)
```

# Simulate Data

Simulate data from Equation: 

$$Y_{ij}=\beta_0+\beta_1t_{ij}+b_{i}+e_{ij}, ~~~ i=1,\ldots,100; j=1,..12,$$

where $(\beta_0,\beta_1)^T=(0.25,-0.50)^T$;
$b_{i} \overset{iid}{\sim}\text{N}(0,0.5)$;
$e_{ij} \overset{iid}{\sim}\text{N}(0,4)$

```{r simdata}

# Generate Data
set.seed(04142022)
n<-100                           # total subjects
nis<-rep(12,n)                   # repeated measure per subject (balanced)
id<-rep(1:n,nis)                 # id = (1,1,..,1,2,2,...,2,...,n,n,...,n)
N<-length(id)                    # total observations


# Fixed Effects
t<-rep(0,N)
for (i in 1:n) t[id==i]<-0:(nis[i]-1)# time variable
X<-cbind(1,t)                        # design matrix 
truebeta<-beta<-c(.25,-.5)           # true beta 
p<-ncol(X)                           # number of parameters

# Random effect variance
truesigmab<-sigmab<-0.5
trueB<-b<-rnorm(n,sd=sqrt(truesigmab))
B<-rep(b,nis)


# Error variance
truesigmae<-sigmae<-4

# Outcome Model
mu<-X%*%beta+B
y<-rnorm(N,mu,sqrt(sigmae))                  
```


# Frequentist Approach
```{r freq}
# Frequentist approach
m <- lmer(y~t+(1|id))
summary(m)
```

# Nimble Code

- $\boldsymbol{b}=(b_1,b_2,\ldots,b_{100})^T$ is a $100\times1$ vector

- ID variable is a $1200\times1$ vector, such as $( \underbrace{1,1,\ldots,1}_{12},\underbrace{2,2,\ldots,2}_{12},\ldots,\underbrace{100,100,\ldots,100}_{12})_{1200\times1}$

```{r nimble}
#--------#
# Nimble #
#--------#
code <- nimbleCode({
  for(i in 1:N) {
    y[i] ~ dnorm(mu[i],tau=taue)
    mu[i] <- beta0+beta1*t[i]+b[id[i]]
  }
  # Random Effect
  for(j in 1:n){b[j] ~ dnorm(0,tau=taub)}  # Random effects
  # Prior 
  taub  ~ dgamma(0.0001,0.0001)            # Random effect precision 
  taue  ~ dgamma(0.0001,0.0001)            # Error precision
  beta0  ~ dnorm(0, sd = 100)              
  beta1  ~ dnorm(0, sd = 100)
  varb <- 1/taub                           # Random effect variance 
  vare <- 1/taue                           # Error variance
})

constants <- list(N=N, n = n,id=id)        # Note: ID variable
data <- list(y = y, t = t)
inits <- list(beta0=0,beta1=0,b=rep(0,n),taub=1,taue=1)

# nimbleModel: check code specification is correct
model <- nimbleModel(code, constants = constants, data = data, inits = inits)
```

# Configure and Compile Model

```{r config}
modelConf <- configureMCMC(model, print = TRUE)
modelConf$addMonitors("varb","vare") # add monitor random effect/error variance
Cmodel <- compileNimble(model)
```

# Compile MCMC Object

```{r compile}
modelMCMC <- buildMCMC(modelConf)
modelMCMC <- compileNimble(modelMCMC, project = model)
```

# Run MCMC

```{r mcmc}
niter <- 10000     # Number of iterations
burn  <- niter/2   # Number of burn-ins
thin  <- 20        # Thinning interval   
set.seed(1)
samples <- runMCMC(modelMCMC, niter = niter,nburnin = burn, nchains = 1,thin = thin,summary = TRUE)
```

# Extract Sample

```{r sample}
samples$summary  # estimate table
```

# MCMC Diagnostics

```{r diag,eval=FALSE}
# Check MCMC plot
out<-samples$samples[,c("beta0","beta1","taub","varb","taue","vare")]
mcmcplot(out,filename = "output")
```

