---
title: "Random Slope Model"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: 72
---

# Model Specification 

-   See Lecture 24 

-   Observation level: Let $Y_{ij}$ be the outcome of subject $i$
    ($i=1,\ldots,n$) at $j$-th ($j=1,\ldots,n_i$) time:
    $$Y_{ij}=\beta_0+\beta_1t_{ij}+b_{1i}+b_{2i}t_{ij}+e_{ij},$$ where
    $t_{ij}\in \{1,2,\ldots,T\}$. We assume
    $\boldsymbol{b}=(b_{1i},b_{2i})^T_{i}\overset{\text{iid}}{\sim}\text{N}_2(0,\Sigma_b)$,
    $e_{ij}\overset{\text{iid}}{\sim}\text{N}(0,\sigma^2_e)$

-   Combining all observations:
    $$\boldsymbol{Y}=\boldsymbol{X}\boldsymbol{\beta}+\boldsymbol{Z}\boldsymbol{b}+\boldsymbol{e},$$
    where
    $\boldsymbol{Y}=(Y_{11},Y_{12},\ldots,Y_{1n_1},\ldots,Y_{n1},Y_{n2},\ldots,Y_{nn_n})_{N\times1}^T$
    and $\boldsymbol{b}=(b_{11},b_{21},b_{12},b_{22},\ldots,b_{1n},b_{2n})_{2n\times1}^T$

-   If we write out the equation above: $$\begin{bmatrix}
    Y_{11}\\
    Y_{12}\\
    \vdots\\
    Y_{1n_1}\\
    \vdots\\
    Y_{n1}\\
    Y_{n2}\\
    \vdots\\
    Y_{nn_n}
    \end{bmatrix}_{N\times1}=\begin{bmatrix}
    1 & t_{11}\\
    1 &t_{12}\\
    \vdots &\vdots\\
    1 &t_{1n_1}\\
    \vdots &\vdots\\
    1 &t_{n1}\\
    1 &t_{n2}\\
    \vdots &\vdots\\
    1 & t_{nn_n}
    \end{bmatrix}_{N\times2}\begin{bmatrix} \beta_0  \\ \beta_1 \end{bmatrix}_{2\times1}+\begin{bmatrix}
    b_{11}\\
    b_{11}\\
    \vdots\\
    b_{11}\\
    \vdots\\
    b_{1n}\\
    b_{1n}\\
    \vdots\\
    b_{1n}
    \end{bmatrix}_{N\times1}+\begin{bmatrix}
    b_{21}t_{11}\\
    b_{21}t_{12}\\
    \vdots\\
    b_{21}t_{1n_1}\\
    \vdots\\
    b_{2n}t_{n1}\\
    b_{2n}t_{n2}\\
    \vdots\\
    b_{2n}t_{nn_n}
    \end{bmatrix}_{N\times1}+
    \begin{bmatrix}
    e_{11}\\
    e_{12}\\
    \vdots\\
    e_{1n_1}\\
    \vdots\\
    e_{n1}\\
    e_{n2}\\
    \vdots\\
    e_{nn_n}
    \end{bmatrix}_{N\times1}$$

```{r setup,warning=FALSE,message=FALSE}
library(mcmcplots)      
library(nimble)
library(lme4)
library(spam)
```

# Simulate Data

Simulate data from Equation: 

$$Y_{ij}=\beta_0+\beta_1t_{ij}+b_{1i}+b_{2i}t_{ij}+e_{ij}, ~~~ i=1,\ldots,100; j=1,..12,$$

where $(\beta_0,\beta_1)^T=(0.25,-0.50)^T$;
$\boldsymbol{b}_{i}=(b_{1i},b_{2i})^T \overset{iid}{\sim}\text{N}_2\left(\begin{bmatrix}0\\0\end{bmatrix},\begin{bmatrix}0.50 & 0.10 \\ 0.10 & 0.15\end{bmatrix}\right)$;
$e_{ij} \overset{iid}{\sim}\text{N}(0,4)$

```{r simdata}
set.seed(04142022)
n<-100                           # total subjects
nis<-rep(12,n)                   # repeated measure per subject (balanced)
id<-rep(1:n,nis)                 # id
N<-length(id)                    # total observations

# Fixed Effects
t<-rep(0,N)
for (i in 1:n) t[id==i]<-0:(nis[i]-1)# time variable: t=0,1,...,11
X<-cbind(1,t)                   # design matrix 
truebeta<-beta<-c(.25,-.5)      # true beta 
p<-ncol(X)                      # number of parameters


# Random Effects
truesigmab<-sigmab<-matrix(c(.50,.10,
                             .10,.15),2,2,byrow = T) # Allow cor. b/t random interept/slope
trueB<-rmvnorm(n,sigma=truesigmab)

trueb1<-b1<-trueB[,1]           # true random intercept 
trueb2<-b2<-trueB[,2]           # true random slope     

B1<-rep(b1,nis)
B2<-rep(b2,nis)

# Precision Error
truesigmae<-sigmae<-4

# Outcome
mu<-X%*%beta+B1+B2*t
y<-rnorm(N,mu,sqrt(sigmae))                
```


# Frequentist Approach
```{r freq}
# Frequentist approach
m <- lmer(y~t+(1+t|id))
summary(m)
```

# Nimble Code

- $\boldsymbol{b}_1=(b_{11},b_{12},\ldots,b_{1,100})^T$ is a $100\times1$ vector for random intercept

- $\boldsymbol{b}_2=(b_{21},b_{22},\ldots,b_{2,100})^T$ is a $100\times1$ vector for random slope

- ID variable is a $1200\times1$ vector, such as $( \underbrace{1,1,\ldots,1}_{12},\underbrace{2,2,\ldots,2}_{12},\ldots,\underbrace{100,100,\ldots,100}_{12})_{1200\times1}$

## Assume Random Intercept and Slope are Independent
```{r nimble0, eval=FALSE}
#--------#
# Nimble #
#--------#
code <- nimbleCode({
  for(i in 1:N) {
    y[i] ~ dnorm(mu[i],tau=taue)
    mu[i] <- beta0+beta1*t[i]+b1[id[i]]+b2[id[i]]*t[i]
  }
  # Random effect
  # update b1i ~ N(0,1/taub1)
  for(j in 1:n){b1[j] ~ dnorm(0,tau=taub1)}  # Random intercept
  # update b2i ~ N(0,1/taub2)
  for(j in 1:n){b2[j] ~ dnorm(0,tau=taub2)}  # Random slope
  # Prior 
  taub1  ~ dgamma(0.0001,0.0001)            # Random intercept precision 
  varb1  <- 1/taub1
  taub2  ~ dgamma(0.0001,0.0001)            # Random slope precision 
  varb2  <- 1/taub2
  beta0  ~ dnorm(0, sd = 100)
  beta1  ~ dnorm(0, sd = 100)
  taue ~ dgamma(0.001,0.001)
  vare <- 1/taue         
})
constants <- list(N=N, n = n,id=id)
data <- list(y = y, t = t)
inits <- list(beta0=0,beta1=0,b1=rep(0,n),b2<-rep(0,n),taub1=1,taub2=1,taue=1)

# nimbleModel: check code specification is correct
model <- nimbleModel(code, constants = constants, data = data, inits = inits)
```

## Assume Random Intercept and Slope are Dependent
```{r nimble}
#--------#
# Nimble #
#--------#
code <- nimbleCode({
  for(i in 1:N) {
    y[i] ~ dnorm(mu[i],tau=taue)
    mu[i] <- beta0+beta1*t[i]+b1[id[i]]+b2[id[i]]*t[i]
  }
  # Random effect
  # update bi=(b1i,b2i)^T ~ MVNORM(0,covb)
  for (j in 1:n) {
    B[j,1:2] ~ dmnorm(B0[1:2],cov=covb[1:2,1:2]) # N(0,sigmab) 
    b1[j]<-B[j,1]
    b2[j]<-B[j,2]}
  # Prior 
  covb[1:2,1:2] ~ dinvwish(C0[1:2,1:2],k)        # IW(diag(2),3)    
  beta0  ~ dnorm(0, sd = 100)
  beta1  ~ dnorm(0, sd = 100)
  taue ~ dgamma(0.001,0.001)
  vare <- 1/taue         
})
constants <- list(N=N, n = n,id=id, B0=rep(0,2), C0=diag(2),k=3)
data <- list(y = y, t = t)
inits <- list(beta0=0,beta1=0,covb=diag(2),B=matrix(0,n,2),taue=1)

# nimbleModel: check code specification is correct
model <- nimbleModel(code, constants = constants, data = data, inits = inits)
```

# Configure and Compile Model

```{r config}
modelConf <- configureMCMC(model, print = TRUE)
modelConf$addMonitors("vare") # add monitor error variance
Cmodel <- compileNimble(model)
```

# Compile MCMC Object

```{r compile}
modelMCMC <- buildMCMC(modelConf)
modelMCMC <- compileNimble(modelMCMC, project = model)
```

# Run MCMC

```{r mcmc}
niter <- 20000     # Number of iterations
burn  <- niter/2   # Number of burn-ins
thin  <- 20        # Thinning interval   
set.seed(1)
samples <- runMCMC(modelMCMC, niter = niter,nburnin = burn, nchains = 1,thin = thin,summary = TRUE)
```

# Extract Sample

```{r sample}
samples$summary  # estimate table
```

# MCMC Diagnostics

```{r diag,eval=FALSE}
# Check MCMC plot
out<-samples$samples[,c("beta0","beta1",
                        "covb[1, 1]","covb[2, 2]","vare","taue")]
mcmcplot(out,filename = "output")
```

